How did the introduction of Stemmed Points affect Mobility and Site Occupation during the Late Pleistocene in Korea?

### 2020-12-07 analysis/paper/paper.Rmd

Previous work on this technological change in Korea has focussed on issues of the transition from the Early to the Late Paleolithic and modern human dispersals [@norton2009evolution; @bae2010origin]. To our knowledge, the implications of the appearance of stemmed points for forager mobility and site use have not been explored in detail. Here, we attempt to identify patterns of mobility through analysis of the stone artefact assemblages using Human Behavioural Ecology (HBE) models. Drawing on established relationships between lithic technology and forager mobility [@shott1986settlement; @Hiscock_1994; @kuhn1994formal; @kuhn2004upper; @kuhn2015artifacts; @kuhn2016moves; @capriles2018], we hypothesize that the appearance of the new hunting tool might reflect a preference for more portable and efficient technologies, that are part of a broader strategy of moving frequently and further, possibly as an adaptation to environmental or population changes.

@@

Studying the raw material consumption patterns is based on an assumption that stone raw materials were transported and consumed to optimize mobility and minimize risk of not having artefacts when they were needed [@andrefsky1994raw; @brantingham2000raw; @brantingham2003neutral; @yue2020lithic]. The representation of raw materials in an assemblage can be an indicator of site occupation patterns. For example, a large variability of raw materials in a site can suggest long-term settlement, while varied and exotic raw materials can indicate higher mobility [@valde2020]. Here we want to know the specific raw materials that relate to the appearance of a new technology, and understanding site occupation patterns by examining changes in the proportions of raw material types in the assemblages.

@@

There is a debate about whether past population growth has played a role in technological change [@shennan2001demography; @kline2010population]. @shennan2009patter observed a correlation between population decrease and technological simplification in their study of  lithic arrowheads from the southern Scandinavian Mesolithic. They found that the lithics became less complex, and point shape changed as the population level dropped. On the other hand, @collard2013population argue that regional archaeological data do not support the population-driven models [@buchanan2016drivers]. For example, @buchanan2016drivers concluded that changes in point types from >13,000 years ago to 400 years ago in Texas were related to environmental risk, and not dependent on population size. We explored SPD to test for correlations between population, environmental risk and technological change. We used the R package ‘rcarbon’ [@cremaanalysing] to perform statistical analyses and conduct model testing using Monte Carlo methods. We generated exponential, uniform, and linear models and evaluated the model goodness-of-fit by applying Akaike’s Information Criterion (AIC) as a selection criterion [@sakamoto1986akaike; @bevan2017holocene; @riris2019widespread].

>>>>

Previous work on this technological change in Korea has focussed on issues of the transition from the Early to the Late Paleolithic and modern human dispersals [@norton2009evolution; @bae2010origin]. To our knowledge, the implications of the appearance of stemmed points for forager mobility and site use have not been explored in detail. Here, we attempt to identify patterns of mobility through analysis of the stone artefact assemblages using Human Behavioural Ecology (HBE) models. Drawing on established relationships between lithic technology and forager mobility [@shott1986settlement; @Hiscock_1994; @kuhn1994formal; @kuhn2004upper; @kuhn2015artifacts; @kuhn2016moves; @capriles2018mobility], we hypothesize that the appearance of the new hunting tool might reflect a preference for more portable and efficient technologies, that are part of a broader strategy of moving frequently and further, possibly as an adaptation to environmental or population changes.

@@

Studying the raw material consumption patterns is based on an assumption that stone raw materials were transported and consumed to optimize mobility and minimize risk of not having artefacts when they were needed [@andrefsky1994raw; @brantingham2000raw; @brantingham2003neutral; @yue2020lithic]. The representation of raw materials in an assemblage can be an indicator of site occupation patterns. For example, a large variability of raw materials in a site can suggest long-term settlement, while varied and exotic raw materials can indicate higher mobility [@valde2020models]. Here we want to know the specific raw materials that relate to the appearance of a new technology, and understanding site occupation patterns by examining changes in the proportions of raw material types in the assemblages.

@@

There is a debate about whether past population growth has played a role in technological change [@shennan2001demography; @kline2010population]. @shennan2009pattern observed a correlation between population decrease and technological simplification in their study of  lithic arrowheads from the southern Scandinavian Mesolithic. They found that the lithics became less complex, and point shape changed as the population level dropped. On the other hand, @collard2013population argue that regional archaeological data do not support the population-driven models [@buchanan2016drivers]. For example, @buchanan2016drivers concluded that changes in point types from >13,000 years ago to 400 years ago in Texas were related to environmental risk, and not dependent on population size. We explored SPD to test for correlations between population, environmental risk and technological change. We used the R package ‘rcarbon’ [@cremaanalysing] to perform statistical analyses and conduct model testing using Monte Carlo methods. We generated exponential, uniform, and linear models and evaluated the model goodness-of-fit by applying Akaike’s Information Criterion (AIC) as a selection criterion [@sakamoto1986akaike; @bevan2017holocene; @riris2019widespread].
